From Parametricity to Conservation Laws, via Noether's Theorem

### 2013-07-11 mechanics/paper.tex
Nearly there

@@

The use of invariance under change to derive useful consequences is a
technique much older than programming languages. In physics,
\emph{Noether's theorem} \cite{noether} provides a general way to
derive conservation laws for physical systems from invariance
properties. Informally speaking, Noether's theorem states that every
continuous symmetry of a physical system implies the existence of a
corresponding conserved quantity. A conserved quantity is a function
of the state of the system that is constant for all time. We
illustrate the use of Noether's theorem with a simple example of a
classical system consisting of two particles of equal mass $m$
connected by a spring with coefficient $k$, such that the particles
are constrained to move one-dimensionally.

@@

From the Lagrangian (\ref{eq:coupled-spring-lagrangian}), using the
\emph{principle of stationary action}, we can derive the following two
equations of motion for this system. (We describe the principle of
stationary action, and the process for deriving the equations of
motion fully in Section~\ref{sec:conservation-laws-from-symmetry}.)
For this system, the equations of motion are a pair of ordinary
differential equations (ODEs) that describe how the positions and
velocities of the particles evolve over time (the second deivative
with respect to time, $\ddot{x_i}$, denotes the acceleration of the
$i$th particle):
\begin{mathpar}
  m\ddot{x_1} = k(x_1 - x_2)

  m\ddot{x_2} = k(x_1 - x_2)
\end{mathpar}
We could now proceed to solve these ODEs to further analyse the
behaviour of this system. However, Noether's theorem gives us a
powerful way of gaining insight into properties of these ODEs ``for
free'', without necessarily having to find solutions to them. This is
accomplished by finding invariance properties of the Lagrangian.

@@

In this paper, we answer this question positively. We show that it is
indeed the case that we can use a generalised version of Reynolds'
theory of relational parametricity to prove the geometric invariance
properties required for the hypotheses of Noether's theorem. We
construct an extension of System F$\omega$ suitable for writing
invariant Lagrangians, with a relationally parametric model that
allows us to prove invariance properties as free theorems. In our
system, the Lagrangian (\ref{eq:coupled-spring-lagrangian}) we gave
above for describing a system of two particles coupled by a spring
will have the type:
\begin{displaymath}
  \begin{array}{@{}l}
  \forall y \mathord: \mathsf{T}(1).\\
  \quad C^\infty(\mathbb{R}\langle 1, 0 \rangle \times \mathbb{R}\langle 1, y \rangle \times \mathbb{R}\langle 1, y \rangle \times \mathbb{R}\langle 1, 0 \rangle \times \mathbb{R}\langle 1, 0 \rangle, \mathbb{R}\langle 1, 0 \rangle)
\end{array}
\end{displaymath}
We explain this type by breaking it down from left to right. The
quantifier $\forall x \mathord: \mathsf{T}(1)$ indicates that we are
quantifying over all translations $y$ in one-dimensional space, just
as we did implicity in Equation
(\ref{eq:coupled-spring-translation-invariance}). The notation
$C^\infty(-,-)$ denotes the type of \emph{smooth} functions between
spaces. A smooth function is a function that is infinitely
differentiable. All the functions that we wish to use as Lagrangians
must be smooth (at least up to second order) for the theory to
work. The type $\mathbb{R}\langle g, x \rangle$ denotes real numbers
that vary with some linear transformation $g$ and translation $x$. In
this case, all the linear transformations are $1$, indicating no
transformation. Likewise, when the translation component is $0$, there
is no translation. Thus a value of type $\mathbb{R}\langle 1, 0
\rangle$ is invariant, but a value of type $\mathbb{R}\langle 1, y
\rangle$ varies with the translation $y$. In the smooth function type
above, in domain space consists of vectors of five real numbers: the
time, the two particle positions, which vary with the translation $y$,
and the two particle velocities.

@@

Our major contributions are threefold:
\begin{enumerate}
\item We reformulate the type system for geometric invariance of
  Atkey, Johann and Kennedy \cite{atkey13abstraction} as an extension
  of System F$\omega$. This shows that the special status of the
  group-indexed types in Atkey \emph{et al.}'s work can be
  incorporated into a standard framework for presenting indexed types.
\item We present a relationally parametric semantics of our extension
  of System F$\omega$. In this model, following Atkey's model for
  basic System F$\omega$ \cite{atkey12relational}, each kind is
  interpreted as a reflexive graph. Our key technical contribution is
  to note that the many geometric groups that we wish to incorporate
  are expressible as groupoids, and hence as reflexive graphs. We
  therefore show that Reynolds' theory of relational parametricity,
  suitably generalised to indexed types by considering reflexive
  graphs, already accomodates geometric invariance properties.
\item Finally, we connect the free theorems derivable from our
  relationally parametric semantics with Noether's theorem for
  deriving conservation laws. We present many examples of invariant
  Lagrangians expressible in our type system, along with the
  consequent conservation laws for physical systems. In each case, the
  necessary invariance property is derived from the type we assign to
  the Lagrangian.
\end{enumerate}

@@

\begin{example}
  We have already seen an example of a Lagrangian, in the
  introduction, describing a system of two particles coupled by a
  spring. Another example of a Lagrangian describes a system
  consisting of a single particle of mass $m$ acted upon by a constant
  downward gravitational field with gravitational acceleration\footnote{is this the right word?} $g$:
  \begin{equation}\label{eq:simple-g-lagrangian}
    L(t,x,y,\dot{x},\dot{y}) = \frac{1}{2}m(\dot{x}^2 + \dot{y}^2) - mgy
  \end{equation}
  Again, the left summand represents the kinetic energy of the system
  in terms of the mass and horizontal and vertical velocities of the
  particle. The right summand represents the potential energy of the
  system due to the gravitational field's strength determined by the
  particle's distance from a fixed baseline at $y = 0$.
\end{example}

@@

The notation in the Euler-Lagrange equation, while standard, is not
exactly clear, especially to a programming language theorist used to a
more careful treatment of variables. The two partial derivatives
$\frac{\partial L}{\partial \dot{q_i}}$ and $\frac{\partial
  L}{\partial q_i}$ take partial derivatives of $L$ with respect to
$\dot{q_i}$ and $q_i$ \emph{as variables}. The derivative with respect
to time, $\frac{d}{dt}$ then treats $\dot{q_i}$ as a function of time,
hence the appearance of the second derivatives $\ddot{q_i}$ in the
equations of motion. Sussman and XXX \cite{sussman} criticise the
standard presentation of Lagrangian mechanics for this kind of
confusing notation, and define a clearer notation based on the Scheme
programming language. We elect to stick with the standard notation as
used in physics textbooks just for reasons of familiarity.

\begin{example}
  Computing the Euler-Lagrange equations for the Lagrangian
  (\ref{eq:simple-g-lagrangian}), above, yields a pair of ODEs, for
  the $x$ and $y$ coordinates of the single particle of the system:
  \begin{mathpar}
    m\ddot{x} = 0

    m\ddot{y} = mg
  \end{mathpar}
  Note that both these ODEs have the form $F = m\ddot{x}$, which is
  exactly Newton's second law. Lagrangian mechanics allows, from the
  principle of stationary action, the derivation of Newton's second
  law, rather than postulating it as an axiom.
\end{example}

@@

This finishes our short introduction to Lagrangian mechanics and
Noether's theorem. We have shown how Noether's theorem can be used to
derive conservation laws from invariance properties. We now go on to
show how suitable invariance properties can be derived from types. In
Section~\ref{sec:examples}, we give examples of invariant Lagrangian's
and their consequent conservation laws.

@@

The first part here will be at least partly taken from Atkey
2012. Work out a way of emphasising that we do not need to do anything
to extend the basic underlying semantic structures involved: reflexive
graphs are enough. This presentation differs from Atkey 2012 in that
we emphasise the reflexive graph structure more, in order to be able
to better expose the connection with the groupoid kinds in the section
below.

\subsection{Relational Parametricity for System F$\omega$}
\label{sec:refl-graphs-for-fomega}

Reflexive graphs give us a way of capturing the higher-dimensional
structure of types in a

Relational parametricity (Reynolds, BFSS), reflexive graphs
(Robinson/Rosolini, Hasegawa, me).

Explain the semantic setting. Explain the ``property'' thing.

\paragraph{Syntax of System F$\omega$} Present the basic syntax of the
kinds, types and terms of (predicative) System F$\omega$.

Small kinds. We will extend the kinds, types and terms later to
encompass the necessary geometrical constructs for classical
mechanics.

\begin{displaymath}
  \kappa ::= * \sepbar \kappa_1 \to \kappa_2 \sepbar \kappa_1 \times \kappa_2 \sepbar \cdots
\end{displaymath}

\begin{mathpar}
  \inferrule*
  {\kappa_1~\mathrm{small} \\ \kappa_2~\mathrm{small}}
  {\kappa_1 \to \kappa_2~\mathrm{small}}

  \inferrule*
  {\kappa_1~\mathrm{small} \\ \kappa_2~\mathrm{small}}
  {\kappa_1 \times \kappa_2~\mathrm{small}}
\end{mathpar}

Type equality... $\Theta \vdash A \equiv B : \kappa$

@@

@@

The interpretation of kinds as reflexive graphs, and small kinds as
small reflexive graphs is a key property of our semantics that we will
maintain as we add additional kinds in
Sections~\ref{sec:discrete-kinds} and~\ref{sec:groupoid-kinds},
below. We state this as Property~\ref{property:semantic-kinds}:

@@

Product kinds are interpreted by taking the product of their
interpretations as reflexive graphs:
\begin{displaymath}
  \begin{array}{@{}l}
    \sem{\kappa_1 \times \kappa_2} = \\
    \begin{array}[t]{l@{}l}
      (& \semKU{\kappa_1} \times \semKU{\kappa_2}, \\
      & ((o_1,o_2),(o_1',o_2')) \mapsto \semKR{\kappa_1}(o_1,o_1') \times \semKR{\kappa_2}(o_2,o_2'), \\
      & (o_1,o_2) \mapsto (\semKI{\kappa_1}(o_1), \semKI{\kappa_2}(o_2)))
    \end{array}
  \end{array}
\end{displaymath}
The objects of the interpretation of a product kind is the product of
the underlying collections of objects of the two parts, and the
relational component is simply the product of the relational
components. This naturally leads to the identity component being
defined as the tuple of the identity components of the two
parts. Again, by assumption that our collection of small sets is
closed under products, if $\sem{\kappa_1}$ and $\sem{\kappa_2}$ are
small reflexive graphs, then so is $\sem{\kappa_1 \times
  \kappa_2}$. Product kinds generalise to the interpretation of
kinding contexts $\Theta = \alpha_1 : \kappa_1, ..., \alpha_n :
\kappa_n$, which are interpreted as the product of the reflexive graph
interpretations of $\kappa_1, ..., \kappa_n$:
\begin{displaymath}
  \begin{array}{l}
    \sem{\alpha_1 : \kappa_1, ..., \alpha_n : \kappa_n} = \\
    \begin{array}[t]{l@{}l}
      (&\semKU{\kappa_1} \times ... \times \semKU{\kappa_n}, \\
      &(\theta,\theta') \mapsto \semKR{\kappa_1}(\pi_1\theta, \pi_1\theta') \times ... \times \semKR{\kappa_n}(\pi_n\theta, \pi_n\theta'), \\
      &\theta \mapsto (\semKI{\kappa_1}(\pi_1 \theta), ..., \semKI{\kappa_n}(\pi_n \theta)))
    \end{array}
  \end{array}
\end{displaymath}

@@

\paragraph{Interpretation of Terms} We omit the straightforward and
relatively uninteresting interpretation of well-typed terms $\Theta
\sepbar \Gamma \vdash e : A$, and just state that there is a
well-defined function interpreting each well-typed term, with the
property that it takes related environments to related results:

\begin{property}
  For all well-typed terms $\Theta \sepbar \Gamma \vdash e : A$ there
  is a function $\sem{e} \in (\forall \theta \in \semKU{\Theta}.\
  \sem{\Gamma}^f\theta \to \sem{A}^f\theta)$, such that, for all
  $\theta, \theta' \in \semKU{\Theta}$, $\rho \in
  \semKR{\Theta}(\theta,\theta')$, $\gamma \in \sem{\Gamma}^f\theta$
  and $\gamma' \in \sem{\Gamma}^f\theta'$, if $(\gamma, \gamma') \in
  \sem{\Gamma}^r\theta\theta'\rho$ then $(\sem{e}\theta\gamma,
  \sem{e}\theta'\gamma') \in \sem{A}^r\theta\theta'\rho$.  Moreover,
  this interpretation is sound for the $\beta\eta$ equational theory
  of terms.
\end{property}

@@

In Section~\ref{sec:refl-graphs-for-fomega}, we only had a single base
kind: the kind $*$ of proper types, with a specific intepretation as
the reflexive graph of sets and relations. We now describe two
families of base kinds with interpretations that are particular sorts
of reflexive graphs. In this section, we look at discrete kinds; kinds
whose reflexive graph interpretations are such that the reflexive
edges are the only edges between objects. In the following section
(Section~\ref{sec:groupoid-kinds}), we look at groupoid kinds, where
edges are composable and invertible.

Discrete kinds can be seen as the natural way of lifting base types up
to the kind level.

\begin{displaymath}
  \kappa ::= \cdots \sepbar \mathsf{nat}
\end{displaymath}

\begin{displaymath}
  \inferrule*
  { }
  {\mathsf{nat}~\mathrm{small}}
\end{displaymath}

\begin{displaymath}
  \sem{\mathsf{nat}} = (\mathbb{N}, (n_1,n_2) \mapsto \{ * \sepbar n_1 = n_2 \}, n \mapsto *)
\end{displaymath}

\begin{mathpar}
  \inferrule*
  { }
  {\Theta \vdash \mathit{zero} : \mathsf{nat}}

@@

  \inferrule*
  {\Theta \vdash A : \mathsf{nat} \\
    \Theta \vdash B : \kappa \\
    \Theta \vdash C : \kappa \to \kappa}
  {\Theta \vdash \mathit{natrec}~A~B~C : \kappa}
\end{mathpar}

Make the point that any set can be used as the interpretation of a
kind. Also, the utility of singleton types.

@@

There is a full and faithful functor from the category of groupoids to
the category of reflexive graphs, that preserves smallness.

\paragraph{The additive group of integers}

\paragraph{Matrix groups}

GL and O, with group operations

$\mathit{ortho}_n : \mathsf{O}(n) \to \mathsf{GL}(n)$
$\mathit{scale}_n : \mathsf{GL}(1) \to \mathsf{GL}(n)$

\paragraph{Translation groups}

$\mathit{exp} : \mathsf{T}(1) \to \mathsf{GL}(1)$
$\underline{c}\cdot- : \mathsf{Z} \to \mathsf{T}(1)$

\paragraph{Groupoid of Cartesian Spaces}

\subsection{Why not Groupoids Everywhere?}

Explain why we aren't using groupoids as kinds: don't need to, doesn't
directly generalise relational parametricity, see Robinson's
paper. Distance-indexed types.

>>>>

\newcommand{\comp}{\mathrel{>\kern-0.45em>\kern-0.45em>}}

@@

The use of invariance under change to derive useful consequences is a
technique much older than programming languages. In physics,
\emph{Noether's theorem} \cite{noether} provides a general way to
derive conservation laws for physical systems from invariance
properties. Informally speaking, Noether's theorem states that every
continuous symmetry of a physical system implies the existence of a
corresponding conserved quantity. A conserved quantity is a function
of the state of the system that is constant for all time. We
illustrate the use of Noether's theorem with a simple example of a
classical system consisting of two particles of equal mass $m$
connected by a spring with spring constant $k$, such that the
particles are constrained to move one-dimensionally.

% Picture here?

@@

From the Lagrangian (\ref{eq:coupled-spring-lagrangian}), using the
\emph{principle of stationary action}, we can derive the following two
equations of motion for this system. (We describe the principle of
stationary action, and the process for deriving the equations of
motion fully in Section~\ref{sec:conservation-laws-from-symmetry}.)
For this system, the equations of motion are a pair of ordinary
differential equations (ODEs) that describe how the positions and
velocities of the particles evolve over time (the second deivative
with respect to time, $\ddot{x_i}$, denotes the acceleration of the
$i$th particle):
\begin{mathpar}
  m\ddot{x_1} = - k(x_1 - x_2)

  m\ddot{x_2} = - k(x_2 - x_1)
\end{mathpar}
We could now proceed to solve these ODEs to further analyse the
behaviour of this system. However, Noether's theorem gives us a
powerful way of gaining insight into properties of these ODEs ``for
free'', without necessarily having to find solutions to them. This is
accomplished by finding invariance properties of the Lagrangian.

@@

In this paper, we answer this question positively. We show that it is
indeed the case that we can use a generalised version of Reynolds'
theory of relational parametricity to prove the geometric invariance
properties required for the hypotheses of Noether's theorem. We
construct an extension of System F$\omega$ suitable for writing
invariant Lagrangians, with a relationally parametric model that
allows us to prove invariance properties as free theorems. In our
system, the Lagrangian (\ref{eq:coupled-spring-lagrangian}) we gave
above for describing a system of two particles coupled by a spring
will have the type:
\begin{displaymath}
  \begin{array}{@{}l}
  \forall y \mathord: \mathsf{T}(1).\\
  \quad C^\infty(\mathbb{R}\langle 1, 0 \rangle \times \mathbb{R}\langle 1, y \rangle \times \mathbb{R}\langle 1, y \rangle \times \mathbb{R}\langle 1, 0 \rangle \times \mathbb{R}\langle 1, 0 \rangle, \mathbb{R}\langle 1, 0 \rangle)
\end{array}
\end{displaymath}
We explain this type by breaking it down from left to right. The
quantifier $\forall x \mathord: \mathsf{T}(1)$ indicates that we are
quantifying over all translations $y$ in one-dimensional space, just
as we did implicity in Equation
(\ref{eq:coupled-spring-translation-invariance}). The notation
$C^\infty(-,-)$ denotes the type of \emph{smooth} functions between
spaces. (A smooth function is a function that is infinitely
differentiable.) All the functions that we wish to use as Lagrangians
must be smooth (at least up to second order) for the theory to
work. The type $\mathbb{R}\langle g, x \rangle$ denotes real numbers
that vary with some linear transformation $g$ and translation $x$. In
this case, all the linear transformations are $1$, indicating no
transformation. Likewise, when the translation component is $0$, there
is no translation. Thus a value of type $\mathbb{R}\langle 1, 0
\rangle$ is invariant, but a value of type $\mathbb{R}\langle 1, y
\rangle$ varies with the translation $y$. In the smooth function type
above, in domain space consists of vectors of five real numbers: the
time, the two particle positions, which vary with the translation $y$,
and the two particle velocities.

@@

Our core contributions are threefold:
\begin{enumerate}
\item We reformulate the type system for geometric invariance of
  Atkey, Johann and Kennedy \cite{atkey13abstraction} as an extension
  of System F$\omega$. This shows that the special status of the
  group-indexed types in Atkey \emph{et al.}'s work can be
  incorporated into a standard framework for presenting indexed types.
\item We present a relationally parametric semantics of our extension
  of System F$\omega$. In this model, following Atkey's model for
  basic System F$\omega$ \cite{atkey12relational}, each kind is
  interpreted as a reflexive graph. Our key technical contribution is
  to note that the many geometric groups that we wish to incorporate
  are expressible as groupoids, and hence as reflexive graphs. We
  therefore show that Reynolds' theory of relational parametricity,
  suitably generalised to indexed types by considering reflexive
  graphs, already accomodates geometric invariance properties.
\item Finally, we connect the free theorems derivable from our
  relationally parametric semantics with Noether's theorem for
  deriving conservation laws. We present many examples of invariant
  Lagrangians expressible in our type system, along with the
  consequent conservation laws for physical systems. In each case, the
  necessary invariance property is derived from the type we assign to
  the Lagrangian.
\end{enumerate}

@@

\begin{example}
  We have already seen an example of a Lagrangian, in the
  introduction, describing a system of two particles coupled by a
  spring. Another example of a Lagrangian describes a system
  consisting of a single particle of mass $m$ acted upon by a constant
  downward gravitational field with gravitational potential $g$:
  \begin{equation}\label{eq:simple-g-lagrangian}
    L(t,x,y,\dot{x},\dot{y}) = \frac{1}{2}m(\dot{x}^2 + \dot{y}^2) - mgy
  \end{equation}
  Again, the left summand represents the kinetic energy of the system
  in terms of the mass and horizontal and vertical velocities of the
  particle. The right summand represents the potential energy of the
  system due to the gravitational field's strength determined by the
  particle's distance from a fixed baseline at $y = 0$.
\end{example}

@@

The notation in the Euler-Lagrange equation, while standard, is not
exactly clear, especially to a programming language theorist used to a
more careful treatment of variables. The two partial derivatives
$\frac{\partial L}{\partial \dot{q_i}}$ and $\frac{\partial
  L}{\partial q_i}$ take partial derivatives of $L$ with respect to
$\dot{q_i}$ and $q_i$ \emph{as variables}. The derivative with respect
to time, $\frac{d}{dt}$ then treats $\dot{q_i}$ as a function of time,
hence the appearance of the second derivatives $\ddot{q_i}$ in the
equations of motion. Sussman and Wisdom \cite{sussman} criticise the
standard presentation of Lagrangian mechanics for this kind of
confusing notation, and define a clearer notation based on the Scheme
programming language. We elect to stick with the standard notation as
used in physics textbooks just for familiarity.

\begin{example}
  Computing the Euler-Lagrange equations for the Lagrangian
  (\ref{eq:simple-g-lagrangian}), above, yields a pair of ODEs, one
  each for the $x$ and $y$ coordinates of the single particle of the
  system:
  \begin{mathpar}
    m\ddot{x} = 0

    m\ddot{y} = -mg
  \end{mathpar}
  Note that both these ODEs have the form $F = m\ddot{x}$, just as in
  Newton's second law. Lagrangian mechanics allows, from the principle
  of stationary action, the derivation of Newton's second law, instead
  of postulating it as an axiom.
\end{example}

@@

This finishes our short introduction to Lagrangian mechanics and
Noether's theorem. We have shown how Noether's theorem can be used to
derive conservation laws from invariance properties. We now go on to
showing how suitable invariance properties can be derived from types,
building towards Section~\ref{sec:types-for-classical-mech}, where we
construct a type system for writing Lagrangians that expresses
invariance properties directly in the types. In
Section~\ref{sec:examples}, we give examples of invariant Lagrangians
and their consequent conservation laws.

@@

We derive invariance properties from types by constructing a
relationally parametric model of System F$\omega$, which we then
extend with suitable kind-, type-, and term-level constants for
writing invariant Lagrangians. Our relationally parametric model is
based on the reflexive graph model presented by
Atkey~\cite{atkey12relational} (based on previous work on reflexive
graph models of relational parametricity by Robinson and
Rosolini~\cite{robinson94} and Hasegawa~\cite{hasegawa94}). In
reflexive graph models of relational parametricity, the kinds of
System F$\omega$ are interpreted as reflexive graphs, while types are
interpreted as morphisms of reflexive graphs. As demonstrated by
Atkey, uniformly interpreting kinds as reflexive graphs allows for a
straightforward interpretation of higher kinds like $* \to *$. In the
present setting, reflexive graphs will allow us to accomodate
arbitrary groupoids as kinds
(Section~\ref{sec:groupoid-kinds}). Groupoid kinds are the key
technical tool that we need to derive the kinds of geometric
invariance properties we need for Noether's theorem.

\subsection{System F$\omega$: Syntax and Relationally Parametric Model}
\label{sec:refl-graphs-for-fomega}

We now present the syntax and relationally parametric semantics of
System F$\omega$. Since we will incrementally extend the syntax of the
system throughout this section and the next, we do not state a final
triple of theorems stating that we have correctly interpreted kinds,
types and terms. Rather, we state the necessary properties of the
interpretation as three separate properties
(Properties~\ref{property:semantic-kinds},
\ref{property:semantic-types}, and~\ref{property:semantic-terms},
below), and maintain them as we extend the system.

We deviate from the standard presentation of System F$\omega$ in that
we distinguish between large and small kinds, and only allow
quantification over small kinds. In particular, the kind of types $*$
is \emph{not} small. The reason for this is that we desire a simple
set-theoretic model. However, requiring that we have a set of sets
closed under large products, in order to interpret quantification over
the kind $*$, would force us to use an intuitionistic metatheory
\cite{pitts}. The basic results of the calculus of variations that we
are relying on, such as the derivation of the Euler-Lagrange
equations, use excluded middle in their proofs, so an intuitionistic
metatheory is not an option. We therefore just assume that we have a
set of ``small'' sets, which includes the real numbers and is closed
under subsets, products, function space and comprehension.

\paragraph{Syntax} The kinds of System F$\omega$ include at least the
base kind of types, $*$, and function and product kinds, as generated
by the following grammar:
\begin{displaymath}
  \kappa ::= * \sepbar \kappa_1 \to \kappa_2 \sepbar \kappa_1 \times \kappa_2 \sepbar \cdots
\end{displaymath}
We will extend the collection of kinds in
Section~\ref{sec:discrete-kinds} with type-level natural numbers, and
in Section~\ref{sec:groupoid-kinds}

We use the judgement $\kappa~\mathrm{small}$ to denote when the kind
$\kappa$ will be interpreted by a small reflexive graph (to be defined
below). The kind $*$ is not small, but the construction of function
and product kinds preserves smallness:
\begin{mathpar}
  \inferrule*
  {\kappa_1~\mathrm{small} \\ \kappa_2~\mathrm{small}}
  {\kappa_1 \to \kappa_2~\mathrm{small}}

  \inferrule*
  {\kappa_1~\mathrm{small} \\ \kappa_2~\mathrm{small}}
  {\kappa_1 \times \kappa_2~\mathrm{small}}
\end{mathpar}
The additional kinds we introduce below will all be small.

The well-kinded types, type equalities, and well-typed terms of System
F$\omega$ are shown in Figures~\ref{fig:types},
\ref{fig:type-equality} and~\ref{fig:terms}, respectively. Note that
typing contexts $\Gamma$ only contain well-kinded types of kind
$*$. Terms also have an equational theory, with $\beta\eta$-laws for
functions, products, and universal quantification, which we omit. Our
presentation of System F$\omega$ is entirely standard (see, e.g.,
Pierce~\cite{pierce01types}), except for the restriction to small
kinds in the formation of universal kinds $\forall
\alpha\mathord:\kappa. A$. As is the case for kinds, we will extend
the types, type equalities and terms in
Sections~\ref{sec:discrete-kinds},~\ref{sec:groupoid-kinds}
and~\ref{sec:types-for-classical-mech}, below.

@@

\begin{figure}
  \centering
  \begin{mathpar}
    \inferrule* [right=$\beta$]
    {\Theta, \alpha : \kappa_1 \vdash A : \kappa_2 \\
      \Theta \vdash B : \kappa_1}
    {\Theta \vdash (\lambda \alpha : \kappa_1. A)~B \equiv A\{B/\alpha\} : \kappa_2}

    \inferrule* [right=$\eta$]
    {\Theta \vdash A : \kappa_1 \to \kappa_2}
    {\Theta \vdash (\lambda \alpha : \kappa_1.A~\alpha) \equiv A : \kappa_1 \to \kappa_2}

    \inferrule* [right=$\beta$]
    {\Theta \vdash A_1 : \kappa_1 \\ \Theta \vdash A_2 : \kappa_2}
    {\Theta \vdash \pi_i~\langle A_1, A_2 \rangle \equiv A_i : \kappa_i}

    \inferrule* [right=$\eta$]
    {\Theta \vdash A : \kappa_1 \times \kappa_2}
    {\Theta \vdash \langle \pi_1 A, \pi_2 A \rangle \equiv A : \kappa_1 \times \kappa_2}
  \end{mathpar}
  {\tiny plus: reflexivity, symmetry, transitivity and congruence}
  \caption{Type equality}
  \label{fig:type-equality}
\end{figure}

@@

The interpretation of kinds as reflexive graphs, and small kinds as
small reflexive graphs is a key property of our semantics that we will
maintain as we add additional kinds in
Sections~\ref{sec:discrete-kinds} and~\ref{sec:groupoid-kinds},
below. We state this as Property~\ref{property:semantic-kinds} of our
semantics:

@@

Product kinds are interpreted by taking the product of their
interpretations as reflexive graphs:
\begin{displaymath}
  \begin{array}{@{}l}
    \sem{\kappa_1 \times \kappa_2} = \\
    \begin{array}[t]{l@{}l}
      (& \semKU{\kappa_1} \times \semKU{\kappa_2}, \\
      & ((o_1,o_2),(o_1',o_2')) \mapsto \semKR{\kappa_1}(o_1,o_1') \times \semKR{\kappa_2}(o_2,o_2'), \\
      & (o_1,o_2) \mapsto (\semKI{\kappa_1}(o_1), \semKI{\kappa_2}(o_2)))
    \end{array}
  \end{array}
\end{displaymath}
The collection of objects of the interpretation of a product kind is
the product of the underlying collections of objects of the two parts,
and the relational component is simply the product of the relational
components. This naturally leads to the identity component being
defined as the tuple of the identity components of the two
parts. Again, by assumption that our collection of small sets is
closed under products, if $\sem{\kappa_1}$ and $\sem{\kappa_2}$ are
small reflexive graphs, then so is $\sem{\kappa_1 \times
  \kappa_2}$. Product kinds generalise to the interpretation of
kinding contexts $\Theta = \alpha_1 : \kappa_1, ..., \alpha_n :
\kappa_n$, which are interpreted as the product of the reflexive graph
interpretations of $\kappa_1, ..., \kappa_n$:
\begin{displaymath}
  \begin{array}{l}
    \sem{\alpha_1 : \kappa_1, ..., \alpha_n : \kappa_n} = \\
    \begin{array}[t]{l@{}l}
      (&\semKU{\kappa_1} \times ... \times \semKU{\kappa_n}, \\
      &(\theta,\theta') \mapsto \semKR{\kappa_1}(\pi_1\theta, \pi_1\theta') \times ... \times \semKR{\kappa_n}(\pi_n\theta, \pi_n\theta'), \\
      &\theta \mapsto (\semKI{\kappa_1}(\pi_1 \theta), ..., \semKI{\kappa_n}(\pi_n \theta)))
    \end{array}
  \end{array}
\end{displaymath}

@@

\paragraph{Interpretation of Terms} We omit the straightforward and
relatively uninteresting interpretation of well-typed terms $\Theta
\sepbar \Gamma \vdash e : A$, and just state that there is a
well-defined function interpreting each well-typed term, with the
property that it takes related environments to related results, this
is the fundamental theorem of logical relations for System F$\omega$:

\begin{property}\label{property:semantic-terms}
  For all well-typed terms $\Theta \sepbar \Gamma \vdash e : A$ there
  is a function $\sem{e} \in (\forall \theta \in \semKU{\Theta}.\
  \sem{\Gamma}^f\theta \to \sem{A}^f\theta)$, such that, for all
  $\theta, \theta' \in \semKU{\Theta}$, $\rho \in
  \semKR{\Theta}(\theta,\theta')$, $\gamma \in \sem{\Gamma}^f\theta$
  and $\gamma' \in \sem{\Gamma}^f\theta'$, if $(\gamma, \gamma') \in
  \sem{\Gamma}^r\theta\theta'\rho$ then $(\sem{e}\theta\gamma,
  \sem{e}\theta'\gamma') \in \sem{A}^r\theta\theta'\rho$.  Moreover,
  this interpretation is sound for the $\beta\eta$ equational theory
  of terms.
\end{property}

@@

In Section~\ref{sec:refl-graphs-for-fomega}, we only had a single base
kind: the kind $*$ of proper types, with a specific interpretation as
the reflexive graph of sets and relations. We now describe two
families of base kinds with interpretations that are particular sorts
of reflexive graph. In this section, we look at discrete kinds; kinds
whose reflexive graph interpretations are such that the reflexive
edges are the only edges between objects. In the following section
(Section~\ref{sec:groupoid-kinds}), we look at groupoid kinds, where
edges are composable and invertible.

Discrete kinds can be seen as the natural way of lifting types up to
the kind level. For our purposes, we will only require a kind of
natural numbers, which we will use for constructing $n$-ary vectors of
positions and velocities when we use our type system for writing
Lagrangians in Section~\ref{sec:types-for-classical-mech}. The kind of
natural numbers is small:
\begin{mathpar}
  \kappa ::= \cdots \sepbar \mathsf{nat}

  \inferrule*
  { }
  {\mathsf{nat}~\mathrm{small}}
\end{mathpar}
The interpretation of this kind as a reflexive graph goes as follows:
\begin{displaymath}
  \sem{\mathsf{nat}} = (\mathbb{N}, (n_1,n_2) \mapsto \{ * \sepbar n_1 = n_2 \}, n \mapsto *)
\end{displaymath}
The collection of objects of this reflexive graph is simply the set of
natural numbers. The only edges in the graph are unique edges between
equal numbers. Since we have assumed that our collection of small sets
contains the natural numbers, we have clearly maintained
Property~\ref{property:semantic-kinds} of our semantics.

We add new type-level constructs for the kind $\mathsf{Nat}$ of
natural numbers, representing zero and successor, and a kind-generic
recursion operator:
\begin{mathpar}
  \inferrule*
  { }
  {\Theta \vdash \mathit{zero} : \mathsf{nat}}

@@

  \inferrule*
  {\Theta \vdash A : \mathsf{nat} \\
    \Theta \vdash B : \kappa \\
    \Theta \vdash C : \kappa \to \kappa}
  {\Theta \vdash \mathit{natrec}_\kappa~A~B~C : \kappa}
\end{mathpar}
These three constructs all have the evident interpretations in terms
of the inductive structure of the natural numbers, and satisfy the
following $\beta$-laws, maintaining
Property~\ref{property:semantic-types} of our semantics:
\begin{displaymath}
  \begin{array}{@{}l@{\hspace{0.3em}}c@{\hspace{0.3em}}l}
    \mathit{natrec}_\kappa~\mathit{zero}~B~C & \equiv & B \\
    \mathit{natrec}_\kappa~(\mathit{succ}~A)~B~C & \equiv & C~(\mathit{natrec}_\kappa~A~B~C)
  \end{array}
\end{displaymath}

@@

Discrete kinds are useful for lifting types up to the kind level, but
they do not really exploit the flexibility of reflexive
graphs. Moreover, they do not provide us with the invariance
properties we require to apply Noether's theorem. By considering kinds
whose interpretations are groupoids, we will be able to derive the
invariance properties that we need.

Recall that a groupoid is a category in which all morphisms have
inverses \cite{maclane98categories}. Functors between categories
always preserve isomorphisms, so morphisms of groupoids are just
functors. Every group is a groupoid with one object and a morphism for
every element of the group: composition is the group operation,
inverses are given by the group inverses, and the identity morphism is
given by the group unit. Homomorphisms between groups are in
one-to-one correspondence with functors between the corresponding
groupoids. An example of a groupoid that is not a group is given by
the collection of cartesian spaces (i.e., $\mathbb{R}^n$ for some
$n$), with all diffeomorphisms (smooth invertible functions) as the
morphisms.

Trivially, every groupoid is a reflexive graph, simply by forgetting
the composition and inverses. Likewise, every functor between
groupoids is a morphism of reflexive graphs, by forgetting the
preservation of composition. Therefore, given any groupoid, we can add
a new kind to System F$\omega$ that is interpreted by that groupoid,
and given any functor $f : \sem{\mathsf{G_1}} \to \sem{\mathsf{G_2}}$
between groupoids interpreting kinds $\mathsf{G_1}$ and
$\mathsf{G_2}$, we obtain a type constant of kind $\mathsf{G_1} \to
\mathsf{G_2}$, with interpretation $f$.

\paragraph{Integers} Our first groupoid kind is $\mathsf{Z}$,
interpreted by additive group of integers. We will use this kind to
state the types of the trigonometric functions $\sin$ and $\cos$ in
Section~\ref{sec:types-for-classical-mech-1}, which are periodic with
period $2\pi$. The kind $\mathsf{Z}$ is small:
\begin{mathpar}
  \kappa ::= \cdots \sepbar \mathsf{Z}
